---
title: "T00711122_ML_FINAL"
output: pdf_document
date: "2023-04-14"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

**1.IMPORTING NECESSARY LIBRARIES**
```{r}
library(ggplot2)      #Importing necessary libraries for my classification methods and related procedures.
library(tidyverse)
library(class)
library(caret)
library(e1071)
library(dplyr)
library(rpart)
library(rpart.plot)
library(randomForest)
```

**2.LOADING AIRLINE DATASET**
```{r}
airline <- read.csv("/Users/keerthanasenthilkumar/Downloads/10K_Airline Passenger Satisfaction - V4.csv", header = T, ",")
```

**(2.1) CHECKING THE COLUMN NAMES IN THE DATASET**
```{r}
names(airline)   
#To know the column names of the airline data set.
```

**(2.2) SUMMARY OF THE DATASET**
```{r}
summary(airline)
#This function displays the minimum, maximum, median, mean, and quartiles for each continuous variable.
``` 

**(2.3) STRUCTURE OF THE DATASET**
```{r}
str(airline)
#It displays the number of observations and variables in the dataset,names of the variables and the data type of each variable.
```

**(2.4) CHECKING THE NULL VALUES**
```{r}
sum(is.na(airline)) 
#To check whether airline dataset have any missing or null values.
```

**3.EXPLORATORY DATA ANALYSIS**

**(3.1) BOXPLOT FOR CHECKING THE OUTLIERS IN CONTINUOUS VARIABLES**

```{r}
ggplot(airline, aes(x = Satisfaction.Rating  , y = Flight.Distance)) +
  geom_boxplot() +
  labs(y = "Flight Distance", x = "Satisfaction Rating" ) +
  ggtitle("Boxplot of Flight Distance")
```
**(3.2) REMOVE THE OUTLIERS**
```{r}
identify_outliers <- function(x) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  upper_fence <- q3 + 1.5*iqr
  lower_fence <- q1 - 1.5*iqr
  outlier_indices <- which(x < lower_fence | x > upper_fence)
  return(outlier_indices)
}

outliers <- identify_outliers(airline$Flight.Distance)
if (length(outliers) > 0) {
  cat("Outliers identified in Flight Distance. \n")
  
# Remove the outliers:
  airline <- airline[!airline$Flight.Distance %in% outliers,]
  cat("Outliers removed from the dataset.\n")
} else {
  cat("No outliers identified in Flight Distance.\n")
}
```

**SCALING**
```{r}
#scaling the continuous variable:
airline$Flight.Distance <- scale(airline$Flight.Distance) 
```


**(3.3) BOXPLOT FOR OTHER CONTINUOUS VARIABLES**
```{r}
boxplot(airline$Arrival.Delay.in.Minutes,main="Arrival delay",ylab="Arrival.Delay.in.Minutes")
boxplot(airline$Departure.Delay.in.Minutes,main="departure delay",ylab="Departure.Delay.in.Minutes")
```

**(3.4) REMOVING OUTLIERS**
```{r}
# Identify and remove outliers:
identify_outliers <- function(x) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  upper_fence <- q3 + 1.5*iqr
  lower_fence <- q1 - 1.5*iqr
  outlier_indices <- which(x < lower_fence | x > upper_fence)
  return(outlier_indices)
}

outliers <- identify_outliers(airline$Arrival.Delay.in.Minutes)
if (length(outliers) > 0) {
  cat("Outliers identified in Arrival delay. \n")
  
  # Remove the outliers from the dataset
  airline <- airline[!airline$Arrival.Delay.in.Minutes %in% outliers,]
  cat("Outliers removed from the dataset.\n")
} else {
  cat("No outliers identified in Arrival delay.\n")
}

outliers <- identify_outliers(airline$Departure.Delay.in.Minutes)
if (length(outliers) > 0) {
  cat("Outliers identified in departure delay. \n")
  
# Remove the outliers :
  airline <- airline[!airline$Departure.Delay.in.Minutes %in% outliers,]
  cat("Outliers removed from the dataset.\n")
} else {
  cat("No outliers identified in departure delay.\n")
}
```

**SCALING**
```{r}
airline$Arrival.Delay.in.Minutes <- scale(airline$Arrival.Delay.in.Minutes) 
airline$Departure.Delay.in.Minutes <-scale(airline$Departure.Delay.in.Minutes)
```


**(3.5) DISTRIBUTION PLOT FOR MY TARGET VARIABLE(Satisfaction Rating):**
```{r}
colors <- c("#E69F00", "#56B4E9", "#009E73", "#D55E00", "#0072B2")

ggplot(airline, aes(x = Satisfaction.Rating, fill = Satisfaction.Rating)) +
  geom_bar() +
  scale_fill_manual(values = colors) +
  labs(title = "Count of satisfaction Rating", x = "Satisfaction.Rating", y = "Count")
```
**(3.6) BARCHART FOR OTHER CATEGORICAL VARIABLES (Gender,Customer type,Type of travel,class)**
```{r}
#setting the colors for each group:
colors <- c("#E69F00", "#56B4E9", "#009E73", "#D55E00", "#0072B2")

#creating bar chart for gender variable:
ggplot(airline, aes(x = Gender, fill = Gender)) +
  geom_bar() +
  scale_fill_manual(values = colors) +
  labs(title = "Count of Gender", x = "Gender", y = "Count")

#creating bar chart for variable customer type:
ggplot(airline, aes(x = Customer.Type , fill = Customer.Type)) +
  geom_bar() +
  scale_fill_manual(values = colors) +
  labs(title = "Count of Customer.Type", x = "Customer.Type", y = "Count")

#creating bar chart for variable travel type:
ggplot(airline, aes(x =Type.of.Travel , fill = Type.of.Travel)) +
  geom_bar() +
  scale_fill_manual(values = colors) +
  labs(title = "Count of Type of Travel", x = "Type.of.Travel", y = "Count")

#creating bar chart for variable class:
ggplot(airline, aes(x = Class, fill = Class)) +
  geom_bar() +
  scale_fill_manual(values = colors) +
  labs(title = "Count of Class", x = "Class", y = "Count")
```

**4.CLASSIFICATION MODELS**

```{r}
set.seed(123)     
#It ensures that the random numbers generated in subsequent commands are reproducible.

#converting categorical variable to factor:
airline$Satisfaction.Rating <- as.factor(airline$Satisfaction.Rating)

#splitting dataset into train and test data set:
train_index <- createDataPartition(airline$Satisfaction.Rating, p = 0.6, list = FALSE)
train <- airline[train_index, ]
test <- airline[-train_index, ]     
#The train data set contains 60% of the observations based on the Satisfaction.Rating variable from the original dataset
#it can be used to train the model, while the test data set contains the remaining 40% 
#it will be used to evaluate the performance of the trained model.
```

**(4.1) SUPPORT VECTOR MACHINE**
```{r}
# Training the SVM model:
svm_model <- svm(Satisfaction.Rating ~ ., data = train, kernel = "linear")

# Making predictions on test data:
svm_pred <- predict(svm_model, test)

# Calculating accuracy:
accuracy <- mean(svm_pred == test$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the SVM model on test data:", round(accuracy * 100, 2), "%\n")
```
**STATISTICS OF MODEL**
```{r}
# Create confusion matrix for the model:
confusionMatrix(svm_pred, test$Satisfaction.Rating)
```

**CONFUSION MATRIX FOR SVM**
```{r}
cm <- confusionMatrix(svm_pred, test$Satisfaction.Rating)

# create the confusion matrix plot
cm_plot <- ggplot(data = as.data.frame(cm$table), 
                  aes(x = Prediction, y = Reference, fill = as.numeric(Freq))) + 
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "#E69F18", high = "#56B4E9") +
  theme_minimal() + 
  labs(title = "Confusion Matrix", x = "Prediction", y = "Reference") +
  geom_text(aes(label = Freq), size = 12, fontface = "bold") + 
  scale_x_discrete(expand = c(0, 0.1)) + 
  scale_y_discrete(expand = c(0, 0.1))+
  guides(fill = guide_colorbar(title = "Frequency"))
cm_plot
```
**ACCURACY OF SVM MODEL ON TRAINING DATASET**
```{r}
#Making prediction on train data set:
svm_pred <- predict(svm_model, train)

# Calculating accuracy:
accuracy <- mean(svm_pred == train$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the SVM model on train data:", round(accuracy * 100, 2), "%\n")
```

**(4.2) DECISION TREE**
```{r}
# creating the decision tree:
d_model <- rpart(Satisfaction.Rating ~ ., data = train, method = "class")

# View the decision tree:
rpart.plot(d_model)

# Making predictions on test data set:
d_pred <- predict(d_model, test, type = "class")

# Calculating accuracy:
accuracy <- mean(d_pred == test$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the decision tree model on test data:", round(accuracy * 100, 2), "%\n")
```
**STATISTICS OF MODEL**
```{r}
# Evaluate the performance of model:
confusionMatrix(d_pred, test$Satisfaction.Rating)
```

**CONFUSION MATRIX FOR DECISION TREE**
```{r}
cm <- confusionMatrix(d_pred, test$Satisfaction.Rating)

# create the confusion matrix plot

cm_plot <- ggplot(data = as.data.frame(cm$table), 
                  aes(x = Prediction, y = Reference, fill = as.numeric(Freq))) + 
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "#E69F18", high = "#56B4E9") +
  theme_minimal() + 
  labs(title = "Confusion Matrix", x = "Prediction", y = "Reference") +
  geom_text(aes(label = Freq), size = 12, fontface = "bold") + 
  scale_x_discrete(expand = c(0, 0.1)) + 
  scale_y_discrete(expand = c(0, 0.1))+
  guides(fill = guide_colorbar(title = "Frequency"))
cm_plot
```
**ACCURACY OF DECISION TREE MODEL ON TRAINING DATASET**
```{r}

# Making predictions on train data set:
d_pred <- predict(d_model, train, type = "class")

# Calculating accuracy:
accuracy <- mean(d_pred == train$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the decision tree model on train data:", round(accuracy * 100, 2), "%\n")
```


**(4.3) RANDOM FOREST**
```{r}
# Build the random forest model:
rf_model <- randomForest(Satisfaction.Rating ~ ., data = train, ntree = 100)

# Evaluating the test performance of the model:
rf_pred <- predict(rf_model, test)

# Calculating accuracy:
accuracy <- mean(rf_pred == test$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the random forest model on test data:", round(accuracy * 100, 2), "%\n")
```

**STATISTICS OF MODEL**
```{r}
# Evaluate the performance of the model
confusionMatrix(rf_pred, test$Satisfaction.Rating)
```

**CONFUSION MATRIX FOR RANDOM FOREST**
```{r}
cm <- confusionMatrix(rf_pred, test$Satisfaction.Rating)

# create the confusion matrix plot
cm_plot <- ggplot(data = as.data.frame(cm$table), 
                  aes(x = Prediction, y = Reference, fill = as.numeric(Freq))) + 
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "#E69F18", high = "#56B4E9") +
  theme_minimal() + 
  labs(title = "Confusion Matrix", x = "Prediction", y = "Reference") +
  geom_text(aes(label = Freq), size = 12, fontface = "bold") + 
  scale_x_discrete(expand = c(0, 0.1)) + 
  scale_y_discrete(expand = c(0, 0.1))+
  guides(fill = guide_colorbar(title = "Frequency"))
cm_plot
```
**ACCURACY OF RANDOM FOREST MODEL ON TRAINING DATASET**
```{r}
#Making prediction on train data set:
rf_pred <- predict(rf_model, train)

# Calculating accuracy:
accuracy <- mean(rf_pred == train$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the random forest model on train data:", round(accuracy * 100, 2), "%\n")
```


**(4.4) NAIVE BAYES CLASSIFIER**
```{r}
# Train the Naive Bayes model
nb <- naiveBayes(Satisfaction.Rating ~ ., data = train)

# Make predictions on the test set
pred <- predict(nb, newdata = test[, -which(names(test) == "Satisfaction.Rating")])

# Calculating accuracy:
accuracy <- mean(pred == test$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the naive bayes model on test data:", round(accuracy * 100, 2), "%\n")
```
`

**STATISTICS OF MODEL**
```{r}
# Evaluate the performance of the model
confusionMatrix(pred, test$Satisfaction.Rating)
```

**CONFUSION MATRIX FOR NAIVE BAYES**
```{r}
cm <- confusionMatrix(pred, test$Satisfaction.Rating)

# create the confusion matrix plot

cm_plot <- ggplot(data = as.data.frame(cm$table), 
                  aes(x = Prediction, y = Reference, fill = as.numeric(Freq))) + 
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "#E69F18", high = "#56B4E9") +
  theme_minimal() + 
  labs(title = "Confusion Matrix", x = "Prediction", y = "Reference") +
  geom_text(aes(label = Freq), size = 12, fontface = "bold") + 
  scale_x_discrete(expand = c(0, 0.1)) + 
  scale_y_discrete(expand = c(0, 0.1))+
  guides(fill = guide_colorbar(title = "Frequency"))
cm_plot
```
**ACCURACY OF NAIVE BAYES MODEL ON TRAINING DATASET**
```{r}

# Make predictions on the train set:
pred <- predict(nb, newdata = train[, -which(names(train) == "Satisfaction.Rating")])

# Calculating accuracy:
accuracy <- mean(pred == train$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the naive bayes model on train data:", round(accuracy * 100, 2), "%\n")
```

**(4.5) KNN**
```{r}
#Train the KNN model:
knn_model <- train(Satisfaction.Rating~., data = train, method = "knn")

#Make prediction on test data set:
knn_predict <- predict(knn_model, newdata = test)

# Calculating accuracy:
accuracy <- mean(knn_predict == test$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the knn model on test data:", round(accuracy * 100, 2), "%\n")
```

**STATISTICS OF MODEL**
```{r}
confusionMatrix(knn_predict, test$Satisfaction.Rating)
```

**CONFUSION MATRIX**
```{r}
cm <- confusionMatrix(knn_predict, test$Satisfaction.Rating)

# create the confusion matrix plot

cm_plot <- ggplot(data = as.data.frame(cm$table), 
                  aes(x = Prediction, y = Reference, fill = as.numeric(Freq))) + 
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "#E69F18", high = "#56B4E9") +
  theme_minimal() + 
  labs(title = "Confusion Matrix", x = "Prediction", y = "Reference") +
  geom_text(aes(label = Freq), size = 12, fontface = "bold") + 
  scale_x_discrete(expand = c(0, 0.1)) + 
  scale_y_discrete(expand = c(0, 0.1))+
  guides(fill = guide_colorbar(title = "Frequency"))
cm_plot
```
**ACCURACY OF KNN MODEL ON TRAINING DATASET**
```{r}
knn_predict <- predict(knn_model, newdata = train)

# Calculating accuracy:
accuracy <- mean(knn_predict == train$Satisfaction.Rating)

# Printing accuracy:
cat("Accuracy of the knn model on train data:", round(accuracy * 100, 2), "%\n")
```


**5.COMPARISON OF FIVE CLASSIFICATION MODELS BASED ON TRAINING AND TESTING'S ACCURACY **

```{r}
model_data <- data.frame(Model = c("SVM", "Decision Tree", "Random Forest", "Naive Bayes","KNN"),
                         Training_Accuracy = c(99.98, 78.44, 99.96, 79.28, 89.35 ),
                         Testing_Accuracy = c(99.97, 77.55, 90.38, 78.94, 86))

# Converting the data frame to long format:
model_data_long <- tidyr::gather(model_data, key = "Accuracy_Type", value = "Accuracy", -Model)

# Plotting a clustered bar chart:
ggplot(model_data_long, aes(x = Model, y = Accuracy, fill = Accuracy_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Model", y = "Accuracy", title = "Comparison of Model Accuracies") +
  scale_fill_manual(values = c("blue", "red"), name = "Accuracy Type") +
  theme_minimal()
```
